Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-06 paper/abstract.tex
fixed hyphenation of stage

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We prove the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanize this proof in Twelf. We also implement the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that asymptotically
improve code efficiency by nearly a linear factor and also
facilitates modular expression of staged programs.

>>>>

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first-stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We prove the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanize this proof in Twelf. We also implement the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that asymptotically
improve code efficiency by nearly a linear factor and also
facilitates modular expression of staged programs.

### 2016-01-06 paper/relatedwork.tex
fixed hyphenation of stage

However unlike splitting, where the goal is to emit code where first stage results
are computed once and then reused in multiple invocations of second
stage execution, the second (dynamic) function produced by binding time separation
does not use the results of the first; instead,
%only serves as a preprocessing step for the subsequent
%specialization phase of partial evaluation.
it has access to the first stage inputs and recomputes all required
first stage computations. As noted by Knoblock and
Ruf\,\cite{knoblock96}, it may be possible to modify the program
bifurcation algorithm to cache and reuse the intermediate results, but
this was never attempted.
Alternatively, our algorithm could potentially be used as the basis of a general
bifurcation algorithm in a partial evaluator.

>>>>

However unlike splitting, where the goal is to emit code where first-stage results
are computed once and then reused in multiple invocations of second-stage execution, 
the second (dynamic) function produced by binding time separation
does not use the results of the first; instead,
%only serves as a preprocessing step for the subsequent
%specialization phase of partial evaluation.
it has access to the first-stage inputs and recomputes all required
first-stage computations. As noted by Knoblock and
Ruf\,\cite{knoblock96}, it may be possible to modify the program
bifurcation algorithm to cache and reuse the intermediate results, but
this was never attempted.
Alternatively, our algorithm could potentially be used as the basis of a general
bifurcation algorithm in a partial evaluator.

### 2016-01-06 paper/semantics.tex
cleaning and authors
fixed hyphenation of stage

All three worlds contain unit, product, function, sum and recursive types defined in the usual fashion.
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of worlds \bbtwo\ and \bbonep, respectively, to types at world \bbonem.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgment \mbox{$\steptwo e {e'}$}.
Since first pass evaluation should not reduce stage two terms, this judgment does nothing but traverse $e$ to find {\tt prev} blocks, under which it performs in-place reductions.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no {\tt prev}s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

@@

The statics and dynamics of \lang\ are related by the type safety theorems below, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
Note how the progress theorem for world \bbonem\ states that every well-typed term must take 
either a standard step or a hoisting step.
\begin{abstrsyn}
\begin{theorem} [Progress]
\label{thm:progress} 
\vspace{-0.75em}
\begin{itemize} 
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
\end{theorem}
\begin{theorem} [Preservation]
\label{thm:preservation} 
\vspace{-0.75em}
\begin{itemize} 
\item If $\typesone e A$ and $\lift e {e'}$, \\
	then $\typestwo q B$ and \mbox{$\typesone [\Gamma,\coltwo y B] {e'} A$}.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}
\end{theorem}
\end{abstrsyn}

>>>>

All three worlds contain unit, product, function, sum and recursive types defined in the usual fashion.
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of worlds \bbtwo\ and \bbonep, respectively, as types at world \bbonem.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgment \mbox{$\steptwo e {e'}$}.
Since first pass evaluation should not reduce stage two terms, this judgment does nothing but traverse $e$ to find {\tt prev} blocks, under which it performs in-place reductions.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first-stage subterms (equivalently, no {\tt prev}s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second-stage binders.

@@

The statics and dynamics of \lang\ are related by the type safety theorems below, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second-stage binders under which we are evaluating.
Note how the progress theorem for world \bbonem\ states that every well-typed term must take 
either a standard step or a hoisting step.
\begin{abstrsyn}
\begin{theorem} [Progress]
\label{thm:progress} 
\vspace{-0.75em}
\begin{itemize} 
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
\end{theorem}
\begin{theorem} [Preservation]
\label{thm:preservation} 
\vspace{-0.75em}
\begin{itemize} 
\item If $\typesone e A$ and $\lift e {e'}$, \\
	then $\typestwo q B$ and \mbox{$\typesone [\Gamma,\coltwo y B] {e'} A$}.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}
\end{theorem}
\end{abstrsyn}

### 2016-01-06 paper/splitting.tex
fixed hyphenation of stage

The correctness of \bbonem\ term splitting with respect to hoisting steps is given in \ref{lem:hoistCorrect}.
Because hoisting steps are nothing but rearrangement of second stage code,
this lemma can use the strong requirement of identical combined terms.

@@

Since direct staged term evaluation (\ref{sec:semantics})
reduces all first stage terms to value forms without any remaining stage two work,
the distinction between \bbonem\ and \bbonep\ is unnecessary.
In contrast, when performing program transformations before the first stage inputs are known,
it is valuable to form a clear distinction between ground and mixed terms.
This was similarly observed in prior work seeking to implement self-applicable partial evaluators\,\cite{Mogensen89a,Mogensen89b}.
While this paper assumes that ground annotations already exist in the input,
it may be possible to use binding-time analysis techniques to automatically insert them.
Ground is also similar to the validity mechanism in ML5\,\cite{murphy05}.

>>>>

The correctness of \bbonem\ term splitting with respect to hoisting steps is given in \ref{lem:hoistCorrect}.
Because hoisting steps are nothing but rearrangement of second-stage code,
this lemma can use the strong requirement of identical combined terms.

@@

Since direct staged term evaluation (\ref{sec:semantics})
reduces all first-stage terms to value forms without any remaining stage two work,
the distinction between \bbonem\ and \bbonep\ is unnecessary.
In contrast, when performing program transformations before the first-stage inputs are known,
it is valuable to form a clear distinction between ground and mixed terms.
This was similarly observed in prior work seeking to implement self-applicable partial evaluators\,\cite{Mogensen89a,Mogensen89b}.
While this paper assumes that ground annotations already exist in the input,
it may be possible to use binding-time analysis techniques to automatically insert them.
Ground is also similar to the validity mechanism in ML5\,\cite{murphy05}.
